Relationship between solar wind, $D_{st}$, and plasmasphere trough mass density on one-hour time scales

### 2016-03-18 paper.tex

The plasma trough is the transition region between the high density and low temperature plasmasphere and the lower density and higher energy magnetosphere \citep{Mayr1968ModelMagnetosphereTemperature}.  Inside of the plasmapause, the density is primarily due to ionospheric photoionization; outside of the plasmapuase, the density has contributions from the co-rotating plasmasphere and conveting magnetosphere plasmas.

Although there are well--validated models, as a function of $L$, $MLT$, and geomagnetic activity, of the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}, \cite{LemaireEarthsPlasmasphere}, \cite{Takahashi2006}, and \cite{Takahashi2010}) and plasmapause boundaries (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}), the dependence of the plasma trough on geomagnetic and solar wind conditions has not been extensively studied until the past decade.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 and found that the average ion mass, $M=\rho/n_e$, correlated with geomagnetic activity, with more negative $D_{st}$ values corresponding to slightly higher average ion mass densities.  They also found a relationship between the average ion mass and 1.5- and 3-day averages of $K_p$, with an increasing correlation as $K_p$ was smoothed to longer timescales.  The CRRES measurements used were made in the MLT range of 12:00 and 18:00 and mostly in the plasma trough (between 2-7$R_E$) on an elliptical, nearly equatorial orbit.

\cite{Denton2006} showed a relationship between $D_{st}$ and the field line distribution of plasma density, and specifically along the same field lines as looked at in later papers (6-8$R_E$). They found that the distribution depends on $L$, \note{finish}.

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, \vinote{Connect to harmonic toroidal frequency} a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincided with with significant changes in $\rho_{eq}$ at for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, an enhancment in $\rho_{eq}$ appeared on the same day as minimum $D_{st}$, indicating the possibility of a solar wind control over $\rho_{eq}$.

The results of \cite{Takahashi2006} and \cite{Takahashi2010} are consistent with other results on changes in mass density in the plasma trough region.  \cite{Yao2008} studied the relationship between $D_{st}$ and ion number density for different ion species in different regions (ring current and plasma sheet) and found a general correlation during each of the four storms considered.

In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} and attempt to identify the solar wind processes that drive changes in the plasma trough density.  There are several issues that are addressed in detail: (1) the plasma density measurements are sparse and (2) the dependence on the north/south component of the solar wind magnetic field is expected to be nonlinear. For northward IMF and low solar wind driving of the magnetosphere, the plasmasphere can expand past geosynchronous orbit where measurements of density are made resulting in an increase in observed density that are not due to solar wind driving; for southward IMF, events studies have shown an increase in density associated with strong southward IMF and enhanced magnetospheric activity [\cite{Takahashi2010}; \cite{Yao2008}].  Issue (1) is addressed by considering \note{Finish}.  Issue (2) is addressed by considering epochs and linear correlations independent of the direction of the IMF and also by separation by averages of the northrward and southward IMF.

@@

The dependence of the plasmapause location on geomagnetic activity was observed as early as \note{Cite}.  \cite{Gallagher1988EmpiricalModelPlasmasphere} used measurements from the Retarding Ion Mass Spectrometer onboard the Dynamics Explorer 1 satellite to model the plasmasphere density with a relationship dependent on $L$ shell, altitude, geomagnetic latitude, and observations grouped by a weighted average of $K_p$.  The developed statistical model showed significant improvements when a weighted average of $K_p$ over three days was used to group the observations used in the model.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset associated with \cite{Denton} and used in \cite{Takahashi2010}, with data available from 1980 through 1991; all other parameters are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements \vinote{More explanation?}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hourly range was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple GOES satellites at the same time in the \cite{Denton} dataset, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

In this work, events are defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in the respective panels of Figure~\ref{AllData}.  In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values. \vinote{Not including \req\ events?}

Figure~\ref{ccplot} shows the long-term trends of GOES 6 based $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows using all non-fill hourly values.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 is found to be $0.94$ matching the value found by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991.  The corresponding 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index \vnote{Cite Kyoto} below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmasphere earthward and the second is a possible increase in density due to enhanced earthward magnetospheric convection.  The movement of the plasmapause boundary has been well documented and parameterized, most recently by \cite{Moldwin2002ModelPlasmapause} and \cite{OBrien2003EmpiricalPlasmapause}.

The second is an increase in $\rho_{eq}$ above a threshold of 40~amu/cm$^3$.  It is expected that these increases will typically occur after period of quiet geomagnetic activity that follow a geomagnetic storm.  In this case, the plasma pause boundary may move outward past the location of measurements and the increase will be due to measurements being made inside the plasmapause.  A second possible reason for the increase could be due to the above-mentioned increase in magnetospheric convention.

@@

Our first analysis uses data only from the time interval 1989-1991 that corresponds with the interval used in the analysis in \cite{Takahashi2010}. For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $4\cdot24$ hours were considered before each event and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered on a window of $\pm 12$ hours of the epoch day was computed, and these averaging windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians across events.)  Error bars for each parameter were computed using the standard deviation of the values used in computing the median divided by the square root of the number of values used.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for these events.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is $1.5$ amu/cm$^3$ (from 18 to 19.5~amu/cm$^3$) instead of the value of $\sim 10$ amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for the non-\req\ medians, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of \req\ values as we use all available measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed. \vinote{Alternate: Note that for $D_{st}$ events, we computed median values using all available data, even when \req\ availability was sparse.}

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that all trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6, from 1983-1991, including the magnitude of change in $\rho_{eq}$ across event onset.

>>>>

The plasma trough is a transition region between the high density and low temperature plasmasphere and the lower density and higher energy magnetosphere \citep{Mayr1968ModelMagnetosphereTemperature}.  Inside of the plasmapause, the density is primarily due to ionospheric photoionization; outside of the plasmapuase, the density has contributions from the co-rotating plasmasphere and convecting magnetosphere plasmas.

There are well--validated models of the plasmapause boundary, as a function of $L$, $MLT$, and geomagnetic activity (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}).  Models also exist for the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}, \cite{LemaireEarthsPlasmasphere}, \cite{Takahashi2006}, and \cite{Takahashi2010}), but and dependence of the plasma trough on geomagnetic and solar wind conditions has not been extensively studied until the past decade.

The dependence of the plasmapause location on geomagnetic activity was observed as early as \note{Cite}.  \cite{Gallagher1988EmpiricalModelPlasmasphere} used measurements from the Retarding Ion Mass Spectrometer onboard the Dynamics Explorer 1 satellite to model plasmasphere density using a relationship dependent on $L$ shell, altitude, geomagnetic latitude, and observations grouped by a weighted average of $K_p$.  The developed statistical model showed significant improvements when a weighted average of $K_p$ over three days was used to group the observations used in the model.  \note{Discuss plasma trough relevance}.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 (obervations were made when CRRESS was in an elliptical, near equatorial orbit). They found that the average ion mass, $M=\rho/n_e$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of 12:00 and 18:00 with most observations having L between 5 and 7 $R_E$, corresponding to the plasma trough region.

\cite{Denton2006} found a weak relationship between $D_{st}$ and and $K_p$ and the field line distribution of $\rho$ using observations of the first three torodial Alfven harmonic fequencies for $L$ in the range of 6-8$R_E$.  A more pronounced peak in the distribution along a filed line near the equator was fround for lower $D_{st}$ and larger $K_p$.  The equatorial value of $\rho$ appeared identical for the bins of $D_{st}$ (-142 to -31 nT and -31 to 37 nT) and $K_p$ (1.5 to 3.4 and 3.4 to 5.9) considered .

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, \vinote{Connect to harmonic toroidal frequency} a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.  They found a high correlation between 27 day averages fo F10.7 and 27-day medians of $\rho_{eq}$.

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincided with with significant changes in $\rho_{eq}$ for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, an enhancment in $\rho_{eq}$ appeared on the same day as minimum $D_{st}$.

\cite{Yao2008} studied the relationship between $D_{st}$ and ion number density for low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-sheels (2-14) had a strong correlation (0.88).  Although the correlation in the plasma trough was not caluclated, the data presented show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

The results in the papers discussed indicate that (1) the mass density in the plasma trough region is best correlated with $F10.7$ and (2) there exists a correlation between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} and attempt to identify a statistical characterization of the relationship between geomagnetic activity and mass density.  In addition, we attempt to identify solar wind processes may drive this changes.

There are several issues that are addressed in detail: (1) the plasma density measurements are sparse, (2) The plasma trough density has a high correlation with F10.7, as do geomagnetic and solar wind parameters (3) the dependence on the north/south component of the solar wind magnetic field is expected to be nonlinear. (For northward IMF and low solar wind driving of the magnetosphere, the plasmasphere can expand past geosynchronous orbit where trough estimates of density are made resulting in an increase in observed density that are not due to solar wind driving; for southward IMF, events studies have shown an increase in density associated with strong southward IMF and enhanced magnetospheric activity [\cite{Takahashi2010}; \cite{Yao2008}]).

Issue (1) is addressed by considering in detail the number of observations that are used in epoch averages around the time of two types of evens, decreases in $D_{st}$ below a threshold, and increases in $\rho_{eq}$ above a threshold.  Issue (2) is addressed using various methods to remove or isolate F10.7 depenences.  Issue (3) is addressed by considering epochs and linear correlations independent of the direction of the IMF and also by separation by averages of the northrward and southward IMF.

In addition, we consider the influence of time averaging by comparing results using daily means and meadians and hourly means and medians of solar wind and magentospheric parameters.

@@

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset of \cite{Denton}, which was used in \cite{Takahashi2010}; data are available from 1980 through 1991.  All other parameters used are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements as described in \cite{Takahashi2010}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hour window was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available in the \cite{Denton} dataset from multiple GOES satellites at the same time, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

In this work, events are defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in their respective panels of Figure~\ref{AllData}.  In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values. \vinote{Not including \req\ events?}

Figure~\ref{ccplot} shows the long-term trends of GOES 6 based $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows using all non-fill hourly values.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0 and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index \vnote{Cite Kyoto} below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmasphere earthward and the second is a possible increase in density due to magnetospheric processes \vnote{cite}.

The second type of event is an increase in $\rho_{eq}$ above a threshold of 20~amu/cm$^3$.  It is expected that these increases will typically occur after period of quiet geomagnetic activity that follow a geomagnetic storm.  In this case, the plasmapause boundary may move outward past the measurement location and the increase will be due to measurements being made inside the plasmapause.  A second possible reason for the increase could be due to the above-mentioned increase in due to magnetospheric processes.

@@

Our first analysis uses data only from the time interval 1989-1991, which corresponds with the interval used in the analysis in \cite{Takahashi2010}. For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $4\cdot24$ hours were considered before each event and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch day was computed, and these averaging windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians across events.)  Error bars for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values used.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for these events.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is $1.5$ amu/cm$^3$ (from 18 to 19.5~amu/cm$^3$) instead of the increase of $\sim 10$ amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and error bars.  For the non-\req\ medians, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of \req\ values as we used all available measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed.

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6, which spans 1983-1991 - a small elevation on the day of the event relative to the day before and day after, with averages after the day of the event being lower than that before.

To determine if the magnitude of the density medians on a given day were statistically different than another day, we used a two-population t-test.  For the interval of 1989-1991, corresponding to the top panel of \ref{DailyAveragedDstEvents}, the null hypothesis that any of the medians from days -1, 0, and 1 are different from each other is rejected with a maximimum p value of \vnote{??}.

   Days     p
-1 and 0
-1 and 1
 0 and 1
Results of test on different between means of $\rho_{eq}$ shown in the top panel of \ref{DailyAverage} on days near or on the day of a $D_{st}$ event.  A difference in the means can be rejected with a probability of false rejection of $p$.

   Days     p
-1 and 0
-1 and 1
 0 and 1
Results of test on different between means of $\rho_{eq}$ shown in the bottom panel of \ref{DailyAverage} on days near or on the day of a $D_{st}$ event.
